\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2023-02-27 paper/si.tex
draft s.i.

Section~\ref{section:pairwise_comparison} gives a brief overview of the pairwise
comparison process. The pairwise comparison process consists of three phases;
(1) the mutation phase (2) the game phase and (3) the update phase. In the update
phase an individual adopts the strategy of another individual based on their
``updating payoffs''. In Section~\ref{section:updating_payoffs} we
describe the conventional approach for calculating updating payoffs and we
present our new approaches.

@@

\begin{figure}[!htbp]
    \centering
    \includestandalone[width=\textwidth]{static/pairwise_comparison_process_diagram}
    \caption{\textbf{Pairwise comparison process phases.} \textbf{A) Initialisation} The process begins
    with a finite population where each member is assigned a given strategy.
    Each color represents a different strategy, and the members are labelled by
    letters. \textbf{B) Mutation phase.} An individual is selected (in the
    example individual A) and with a given probability \(\mu\) that individual
    adopts a new strategy. \textbf{C) Game phase.} Individuals are
    selected to interact in a repeated social dilemma with other individuals. We
    demonstrate the case where individuals B and F have been selected to
    interact. They use the reactive strategies \(s_{B} = (y_B, p_B, q_B)\) and
    \(s_{F} = (y_F, p_F, q_F)\) respectively. The opening moves depend on their
    \(y_i\) probability. In turn 1, individual B cooperated,
    thus, F cooperates with a probability \(p_{F}\) in turn 2. On the opposite,
    individual F defected in turn 1, and so B cooperates in the next turn with
    a probability \(q_{B}\). At each turn there is a probability \(\delta\) that a
    subsequent turn will occur, and with a probability (\(1- \delta\)) the interaction ends.
    \textbf{D) Update phase.} At the updating phase two individuals are chosen;
    one serves as the role of the learner and the other one as the 
    role model. In our example C adopts D's strategy with a probability
    \(\rho(\pi_{C}, \pi_{D})\) where \(\pi_{C}, \pi_{D}\) denote the updating
    payoffs of the individuals.}\label{fig:pairwise_phases}
\end{figure}

@@

\section{Updating Payoffs}\label{section:updating_payoffs}

@@

In the following subsections we present several approaches for calculating
the updating payoffs. Initially, we discuss the conventional approach of the
expected payoffs and afterwards we present our new approach.

\subsection{Updating Payoffs based on the expected payoffs}\label{section:perfect_memory}

The expected payoffs are the conventional payoffs used in the updating stage. We
refer to this approach as the perfect memory approach. They are defined as the
mean payoff of an individual in a well-mixed population that engages in an
infinitely repeated games with all other population members. In an infinitely
repeated game the payoff of a reactive strategy can explicitly be calculated
using a markovian approach~\cite{nowak:APC:1989}. Namely, assume two reactive
strategies $s_1\!=\!(y_1, p_1, q_1$) and $s_2\!=\!(y_2,p_2,q_2)$, their play can
be defined as a Markov process with the transition matrix \(M\),

@@

The stationary vector \(\mathbf{v}(s_1,s_2)\),
which is the solution to \(\mathbf{v}(s_1,s_2) M = \mathbf{v}(s_1,s_2)\), combined with the payoff vector
\(\mathbf{u}\), yields the game stage outcome for each strategy,

@@

\subsubsection*{Simulation Results based on Expected Payoffs}

We simulate the evolutionary process, as described in
Algorithm~\ref{algorithm:pairwise_comparison}, when individuals use reactive
strategies and update their strategies based on their expected payoffs. We
performed two independent runs of the process where we differed the benefit of
cooperation \(b\). The results are shown in
Figure~\ref{fig:expected_payoffs_results}. We observed that a higher value of
benefit results in a more cooperative population. For a low benefit the resident
population cooperates on average 52\% of the time. For a high benefit the
average cooperation increases to 98\%. For a low benefit the resident population
consists either of defectors or conditional cooperators, and for a high value of
benefit the population predominantly consists of conditional cooperators.A
conditional cooperator, otherwise known as a generous tit for tat (GTFT)
player, always cooperates if the co-player cooperated ($p\approx1$) and
cooperates with a probability \(q\) if the co-player defected. The conditional
cooperator strategy adopted by the population differs between the two
simulations, namely, a higher value of benefit results in higher values
of \(q\).

\begin{figure}[!htbp]
    \centering 
    \includegraphics[width=.70\textwidth]{static/expected_payoffs_donation_game.pdf}
    \caption{\textbf{Evolutionary dynamics under expected payoffs with
    low (left) and high (right) benefit.} We run the simulations for \(T =
    10^7\) time steps. For each time step we recorded the current resident
    population \((y, p, q)\). Since \(\delta \rightarrow
    1\) we do not report the players' initial cooperation probability \(y\).
    The graphs show how often the resident population chooses each combination
    \((p, q)\) of conditional cooperation probabilities in the subsequent
    rounds. In both cases players update based on their expected payoffs.
    ({\bf Low Benefit}) In
    the case where \(b=3\), the resident population either consists of defectors
    (with \(p \approx q \approx 0\)) or of conditional cooperators for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{3}=0.7\). ({\bf High Benefit}) In the case where
    \(b=10\), the resident population typically applies a strategy for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{10}=0.9\). Parameters: \(N =100, c=1,
    \beta=1\).
    }\label{fig:expected_payoffs_results}
\end{figure}

\subsubsection*{Invasion Analysis based on Expected Payoffs}

We can explicitly calculate the value of \(q\) which evolves. Namely, we will
calculate how easily a single defecting mutant can invade into a resident
population of GTFT players. Let GTFT\(= (1, 1, q)\), ALLD\(= (0, 0, 0)\), and
\(k = 1\). When two GTFT players interact in an infinitely repeated game, their
respective probabilities for each of the four outcomes in the last round
simplify to,

@@

\subsection{Updating Payoffs based on the last last \(m\) round(s) payoffs of \(n\) interaction(s)}

The aim of this work is to explore the effect of the updating payoffs on the
evolution of cooperation. To this end we introduce a new method for calculating
the updating payoffs and we compare the results of the new method to those of
the expected payoffs. In our approach we do not assume that an individual has a
perfect memory. Instead an individual remembers only a limited number of
interactions and limited outcomes from each interaction. In the expected payoffs
case the payoff of a pair depends on the average payoff they received over an
infinite number of turns. In our approach, the payoff of a pair depends on the
average payoffs they received in the last \(m\) turns. Furthermore, in expected
payoffs it is assumed that a player interacts with every member of the
population, whereas in our approach a player has \(n\) interactions.  Here we
present results for four cases; (\(m=1\) and \(n=1\)), (\(m=2\) and \(n=1\)),
(\(m=1\) and \(n=2\)), and (\(m=2\) and \(n=2\)).

\subsubsection{Updating Payoffs based on the Last Round Payoff of One Interaction (\(m=1\) and \(n=1\))}\label{section:m_one_n_one}

In this case an individual updates her/his strategy based on the last payoff
they received against one other member of the population. Initially, we define
the probability that a reactive strategy receives the payoff $u\!\in\!
\mathcal{U}$ in the very last round of the game against another reactive
strategy (Proposition~\ref{proposition:last_round}).

\begin{Prop}\label{proposition:last_round} Consider a repeated game, with
    continuation probability $\delta$, between players with reactive strategies
    $s_1\!=\!(y_1, p_1, q_1$)  and $s_2\!=\!(y_2,p_2,q_2)$ respectively. Then
    the probability that the $s_1$ player receives the payoff $u\!\in\!
    \mathcal{U}$ in the very last round of the game is given by
    $v_{u}(s_1,s_2)$, as given by Equation~(\ref{Eq:LastRound}).

@@

\subsubsection*{Simulation Results based on the Last Round Payoff of One Interaction}

We simulate the evolutionary process given that individuals now use the last
round payoff of one interaction to update their strategies. The results are
shown in Figure~\ref{fig:one_interaction_last_round_payoffs_results}. Similarly
to the results of the expected payoffs, a higher benefit results in a more
cooperative population. However, we note that the cooperation rate of the
evolved populations is lower compared to the case of expected payoffs, and that
the increase of the cooperation rate between the two simulations is smaller. The
residents population in both cases consist of defectors and conditional
cooperators with \(p \approx 1\) and  \(q \approx \frac{1}{2}\). Thus, the
evolved \(q\)s are smaller which is the reason why the population is less
cooperative. In the invasion analysis we demonstrate that in the case of the
last round payoff of one interaction, a conditional player needs to have \(q
\leq \frac{1}{2}\) to avoid being invaded by a defector regardless of the
benefit of cooperation.

\begin{figure}[!htbp]
  \centering 
  \includegraphics[width=.70\textwidth]{static/one_interaction_last_round_donation_game.pdf}
  \caption{\textbf{Evolutionary dynamics under one interaction last round
  payoffs with low (left) and high (right) benefit.} The cooperation rate
  regardless of the benefit of cooperation is less compared to that of the
  expected payoffs. The increase of cooperation between the low benefit case and
  the high benefit case is also smaller. In both cases the resident population
  either consists of defectors (with
  \(p \approx q \approx 0\)) or of conditional cooperators for which \(p \approx
  1\) and \(q \leq \frac{1}{2}\). Parameters: \(N =100, c=1, \beta=1, T=10^{7}\).
  }\label{fig:one_interaction_last_round_payoffs_results}
\end{figure}

\subsubsection*{Invasion Analysis based on the Last Round Payoff of One Interaction}

@@

\subsubsection{Updating Payoffs based on the last round payoff of two interactions (\(m=1\) and \(n=2\))}\label{section:m_one_n_two}

Here we consider the case where both the role model and the learner estimate
their updating payoff after interacting with two members of the population.
There are two stages of matching and there are twenty four possible pairs. In
the first stage the role model and the learner are matched together with a
probability $\frac{1}{N-1}$ or not with a probability $(1 - \frac{1}{N-1})$. If
they were matched together then in the second stage there are only four possible
outcomes; both of them interact with a mutant with a probability
$\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and both interact with a resident with a
probability $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last two possible pairs are
that either of them interacts with a resident whilst the other interacts with a
mutant, and this happens with a probability $\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$.
In the later case, where they were not matched in the first stage, there are
four possible outcomes; both of them interact with a mutant and both interact
with a resident, either of them interacts with a resident whilst the other
interacts with a mutant. For each of the above pairs of the first stage there
are five possible pairs in the second stage; they interact with each other, both
of them interact with a mutant and both interact with a resident, either of them
interacts with a resident whilst the other interacts with a mutant.

@@

\subsubsection*{Simulation Results based on the Last Round Payoff of Two Interactions}

We simulate the evolutionary process when individuals update their strategies
based on the last round payoffs from two interactions with members of the
population, Figure~\ref{fig:one_interaction_last_round_payoffs_results}. For
both a low and a high benefit value the average cooperation rate has increased
compared to the last round against a single other member of the population. In
the case of low benefit the evolved population's cooperation rate is almost the
same as in the case of expected payoffs. However, the cooperation rate does not
increase as much as the expected payoffs for \(b=10\).

\begin{figure}[!htbp]
  \centering 
  \includegraphics[width=.70\textwidth]{static/two_interactions_donation_game.pdf}
  \caption{\textbf{Evolutionary dynamics under two interactions last round
  payoffs with low (left) and high (right) benefit.}
  The cooperation rates for both the simulations have increased and are similar
  to the case of expected payoffs. The increase of the cooperation rate between
  the two cases however is smaller. In both cases the resident population
  either consists of defectors (with
  \(p \approx q \approx 0\)) or of conditional cooperators for which \(p \approx
  1\) and \(q \leq \frac{1}{2}\). Parameters: \(N =100, c=1, \beta=1, T=10^{7}\).
  }\label{fig:two_interactions_results}
\end{figure}

\subsubsection*{Invasion Analysis based on the Last Round Payoff of Two Interactions}

@@

\subsection{Updating Payoffs based on the Last Two Rounds Payoff of One Interaction (\(m=2\) and \(n=1\))}\label{section:m_two_n_one}

In this section we consider the case of \(n=1\) and \(m=2\), thus individuals
update based on their last two rounds payoffs with one member. To this end we
define the probability that a reactive strategy receives the payoffs $u\!\in\!
\mathcal{\tilde{U}}$, for \(\mathcal{\tilde{U}}= \{rrr\dots r, rrr\dots s, \dots,
ppp\dots p\}\), in the last \(m\) rounds against another reactive strategy
(Proposition~\ref{proposition:last_m_rounds}).

@@

\subsubsection*{Simulation Results based on the Last Two Rounds Payoff of One Interaction}

The simulation results for this case are given in
Figure~\ref{fig:two_rounds_results}. Updating on the two rounds payoffs received
against a single opponent result to a lower cooperation rate compare to updating
based on one round against two opponents.

\begin{figure}[!htbp]
  \centering 
  \includegraphics[width=.70\textwidth]{static/two_rounds_donation_game.pdf}
  \caption{\textbf{Evolutionary dynamics under two interactions last round
  payoffs with low (left) and high (right) benefit.}\\Parameters: \(N =100, c=1, \beta=1, T=10 ^7\).
  }\label{fig:two_rounds_results}
\end{figure}

\subsubsection*{Invasion Analysis based on the Last Two Rounds Payoff of One Interaction}

@@

\subsection{Updating Payoffs based on the last two rounds payoff of two interactions (\(m=2\) and \(n=2\))}

The last case we present in this work is the case of \(m = 2\) and \(n = 2\).
That is, an individual considers the last two rounds of payoffs he or she
received against two different members of the population. For this case we
combine the approaches described in sections~\ref{section:m_one_n_two}
and~\ref{section:m_two_n_one}. Here we only present results on the numerical
simulations, Figure~\ref{fig:two_rounds_opponents_results}. The evolved
cooperation rates are between the other cases and the expected payoffs. For a
low benefit we get results similar to the case where one the last round was
considered, however, in the high benefit case the evolved cooperation rate
increases to 91\% which is higher then any of the newly introduced approaches.

\begin{figure}[!htbp]
  \centering 
  \includegraphics[width=.70\textwidth]{static/two_rounds_opponents_donation_game.pdf}
  \caption{\textbf{Evolutionary dynamics under two interactions last round
  payoffs with low (left) and high (right) benefit.} In the cases of low and
  high benefit case the resident population either consists of defectors (with
  \(p \approx q \approx 0\)) or of conditional cooperators for which \(p \approx
  1\) and \(q \leq \frac{1}{2}\). Parameters: \(N =100, c=1, \beta=1\).
  }\label{fig:two_rounds_opponents_results}
\end{figure}

\newpage

\section{Expected and Last Round Updating Payoffs for Memory One Strategies}

So far we have considered the case where members can adopt reactive strategies.
To demonstrate that our results hold for higher memory strategies here we
present results for the expected payoffs, and the last round payoff when members
use memory-one strategies. Memory-one strategies consider the outcome of the
previous round to decide on an action. There are four possible outcome in each
round; \((C, C), (C, D), (D, C), (D, D)\). A memory-one strategy \(s\) can be
written as a five-dimensional vector \(s=(y, p_1, p_2, p_3, p_4)\). The
parameter \(y\) is the probability that the strategy opens with a cooperation
and \(p_1\), \(p_2\), \(p_3\), \(p_4\) are the probabilities that the strategy
cooperates for each of the possible outcomes of the last round.

We perform four separate simulations where we differ the updating payoff and the
benefit of cooperation \(b\). The results for a low value of benefit are given
in Figure~\ref{fig:memory_one_low_benefit}, and for a high benefit in
Figure~\ref{fig:memory_one_high_benefit}.

@@

In this section we evaluate the main result of this work for \(\mu \neq 0\).
Namely, we explore the evolved population when individuals use perfect and
limited updating payoff memory for different values of \(\mu\). We perform five
independent runs of the pairwise process described in
Section~\ref{section:pairwise_comparison}, and at each time step we record the
average player \(\bar{s}=(\bar{y}, \bar{p}, \bar{q})\).

>>>>

Section~\ref{section:pairwise_comparison} gives a brief overview of the pairwise
comparison process. The pairwise comparison process consists of three phases;
(1) the mutation phase (2) the game phase and (3) the update phase. In the
update phase an individual adopts the strategy of another individual based on
their ``updating payoffs''. In the paper we explore the effect of the updating
payoffs on the evolved populations. We used as an example the prisoner's
dilemma. We focused mainly on two approaches for calculating the updating
payoffs which we referred to as the perfect memory and the limited memory
approaches. In Section~\ref{section:perfect_memory} we describe the perfect
memory approach and in Section~\ref{section:limited_memory} we present the
limited memory approach. In the limited memory approach we assume that
individuals update based on their last round payoff against one other member of
the population. The framework can easily be extended to consider more rounds,
more interactions or both. In
Sections~\ref{section:m_one_n_two}-\ref{section:m_two_n_two} we present each of
this extensions and for each a special case.
In Section~\ref{section:simulation_results} we present numerical results of the
pairwise comparison process using the several updating payoff approaches we have
presented. In Sections~\ref{section:memory_one} and~\ref{section:mutation} we
verify the main result of the paper when we no longer assume (i) that
individuals use reactive strategies but instead memory one strategies (ii) low
mutation.

% The aim of our work is to explore the effect
% of the updating payoffs on the evolution of cooperation. To this end we
% introduce a new method for calculating the updating payoffs and we compare the
% results of the new method to those of the expected payoffs. In our approach we
% do not assume that an individual has a perfect memory. Instead an individual
% remembers only a limited number of interactions and limited outcomes from each
% interaction.

@@

\begin{figure}[!htbp]
    \centering
    \includestandalone[width=\textwidth]{static/pairwise_comparison_process_diagram}
    \caption{\textbf{Pairwise comparison process phases.} \textbf{A) Initialisation.} The process begins
    with a finite population where each member is assigned a given strategy.
    Each color represents a different strategy, and the members are labelled by
    letters. \textbf{B) Mutation phase.} An individual is selected (in the
    example individual A) and with a given probability \(\mu\) that individual
    adopts a new strategy. \textbf{C) Game phase.} Individuals are
    selected to interact in a repeated social dilemma with other individuals. We
    demonstrate the case where individuals B and F have been selected to
    interact. They use the reactive strategies \(s_{B} = (y_B, p_B, q_B)\) and
    \(s_{F} = (y_F, p_F, q_F)\) respectively. The opening moves depend on their
    \(y_i\) probability. In turn 1, individual B cooperated,
    thus, F cooperates with a probability \(p_{F}\) in turn 2. On the opposite,
    individual F defected in turn 1, and so B cooperates in the next turn with
    a probability \(q_{B}\). At each turn there is a probability \(\delta\) that a
    subsequent turn will occur, and with a probability (\(1- \delta\)) the interaction ends.
    \textbf{D) Update phase.} At the updating phase two individuals are chosen;
    one serves as the role of the learner and the other one as the 
    role model. In our example C adopts D's strategy with a probability
    \(\rho(\pi_{C}, \pi_{D})\) where \(\pi_{C}, \pi_{D}\) denote the updating
    payoffs of the individuals.}\label{fig:pairwise_phases}
\end{figure}

@@

\subsection{Updating Payoffs}\label{section:updating_payoffs}

@@

In the following sections we present several approaches for calculating the
updating payoffs. Initially, we discuss the conventional approach of the
expected payoffs and afterwards we present our newly introduced approaches.

\section{Updating Payoffs based on the expected payoffs (perfect memory)}\label{section:perfect_memory}

The expected payoffs are the conventional payoffs used in the updating stage. We
refer to this approach as the perfect memory approach. The expected payoffs are
defined as the mean payoff of an individual in a well-mixed population that
engages in an infinitely repeated games with all other population members.
The game between two reactive strategies $s_1\!=\!(y_1, p_1, q_1$) and $s_2\!=\!(y_2,p_2,q_2)$
can be described by a Markovian process~\cite{nowak:APC:1989} with the
transition matrix \(M\),

@@

The stationary vector \(\mathbf{v}(s_1,s_2)\) is the solution to
\(\mathbf{v}(s_1,s_2) M = \mathbf{v}(s_1,s_2)\). In an infinitely repeated game
the game stage outcome for each strategy is given by,

@@

\subsubsection*{Invasion Analysis}

Let's assume that individuals use their expected payoffs at the updating phase.
We can calculate how easily a single defecting mutant (ALLD) can invade into a
resident population of conditional cooperators, otherwise known as generous tit
for tat (GTFT), players. Let GTFT\(= (1, 1, q)\), ALLD\(= (0, 0, 0)\), and \(k =
1\). When two GTFT players interact in an infinitely repeated game, the
stationary distribution \(\mathbf{v}(\text{GTFT}, \text{ALLD})\) 
simplifies to,

@@

This result will become important when we analyse the behaviour that emerges
from the evolutionary process. We will repeat this ``invasion analysis'' for the
rest of the updating payoffs cases.

\section{Updating Payoffs based on the Last Round Payoff of One Interaction (limited memory)}\label{section:limited_memory}

In the expected payoffs case the payoff of a pair depends on the average payoff
they received over an infinite number of turns. In the limited memory payoffs,
the payoff of a pair depends on the average payoffs they received in the last
turn. Furthermore, in expected payoffs it is assumed that a player interacts
with every member of the population, whereas in the limited memory approach a
player has one interaction. Initially, we define the probability that a reactive
strategy receives the payoff $u\!\in\! \mathcal{U}$ in the very last round of
the game against another reactive strategy
(Proposition~\ref{proposition:last_round}).

\begin{Prop}\label{proposition:last_round} Consider a repeated game, with
    continuation probability $\delta$, between players with reactive strategies
    $s_1\!=\!(y_1, p_1, q_1$)  and $s_2\!=\!(y_2,p_2,q_2)$ respectively. Then
    the probability that the $s_1$ player receives the payoff $u\!\in\!
    \mathcal{U}$ in the very last round of the game is given by
    $v_{u}(s_1,s_2)$, as given by Eq.~(\ref{Eq:LastRound}).

@@

\subsubsection*{Invasion Analysis}

@@

\section{Updating Payoffs based on the last round payoff of $n$ interactions}\label{section:m_one_n_two}

In the limited memory payoffs we assume that an individual recalls the last
round they received against one member of the population. This is an edge case.
The framework of the limited memory can be generalised such that an individual
considers \(m\) rounds and of \(n\) interactions. Here we discuss the case that
the update depends on the last round of \(n\) interactions.

At each step of the evolutionary process the role model and the learner now
participate in \(n\) matches. We need to define the probability that for these
matches they are paired with a mutant, with a resident or with each other. We
assume that each pair is unique, so the resident and role model can be matched
together only once at each step. A representation of the process is given in
Figure.

In the case of \(n=1\) there are five possible pairs, however, the number of
possible pairs increases non linearly as we increase the number of possible
interactions. We demonstrate this case for \(n=2\), namely, for when the role
model and the learner have two interactions.

\subsection*{Special case: Last Round Payoff of Two Interactions}

For \(n=2\) there are two stages of matching and there are twenty four possible pairs. In
the first stage the role model and the learner are matched together with a
probability $\frac{1}{N-1}$ or not with a probability $(1 - \frac{1}{N-1})$. If
they were matched together then in the second stage there are only four possible
outcomes; both of them interact with a mutant with a probability
$\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and both interact with a resident with a
probability $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last two possible pairs are
that either of them interacts with a resident whilst the other interacts with a
mutant, and this happens with a probability $\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$.
In the later case, where they were not matched in the first stage, there are
four possible outcomes; both of them interact with a mutant and both interact
with a resident, either of them interacts with a resident whilst the other
interacts with a mutant. For each of the above pairs of the first stage there
are five possible pairs in the second stage; they interact with each other, both
of them interact with a mutant and both interact with a resident, either of them
interacts with a resident whilst the other interacts with a mutant.

@@

\subsubsection*{Invasion Analysis}

In a similar fashion we can calculate the condition for which a population of
GTFT players can not be invaded by an ALLD mutant. Using the new formulation we
obtain the following probabilities \(\bar{x}(u_1, u_2)\) that the payoff of a randomly
chosen GTFT player is \(u_1\) and that the payoff of the ALLD player is \(u_2\),

The ratio of transition probabilities is given by,

@@

\section{Updating Payoffs based on the Last $m$ Rounds Payoffs of One Interaction}\label{section:m_two_n_one}

The second generalised case of the limited memory payoffs we discuss in that of
individuals updating based on their last $m$ rounds payoffs with one member. To
this end we define the probability that a reactive strategy receives the payoffs
$u\!\in\! \mathcal{\tilde{U}}$, for \(\mathcal{\tilde{U}}= \{\underbrace{rrr\dots r}_{m},
\underbrace{rrr\dots s}_{m}, \dots, \underbrace{ppp\dots p}_{m}\}\), in the last \(m\) rounds against another
reactive strategy (Proposition~\ref{proposition:last_m_rounds}).

@@

\subsection*{Special case: Last Round Payoff of Two Interactions}

\subsubsection*{Invasion Analysis}

@@

\section{Updating Payoffs based on the last two rounds payoff of two interactions (\(m=2\) and \(n=2\))}\label{section:m_two_n_two}

Finally a possible extension to the limited memory framework is to consider that
thee number of rounds and number of interactions increase. For this case we need
to consider a combination of the methods we presented in
Section~\ref{section:m_one_n_two} and Section~\ref{section:m_two_n_one}.

\section{Simulation Results on the Pairwise Comparison Process}\label{section:simulation_results}

We simulate the evolutionary process described in
Algorithm~\ref{algorithm:pairwise_comparison} for the different updating
mechanisms we have described in
Sections~\ref{section:perfect_memory}-\ref{section:m_two_n_two}. For each
approach we performed an independent run of the process and for each time step we
recorded the current resident population \((y, p, q)\). The results are shown in
Figure~\ref{fig:expected_payoffs_results}. We observe that in most cases the
resident population consists either of defectors or conditional cooperators. A
conditional cooperator always cooperates if the co-player cooperated
($p\approx1$) and cooperates with a probability \(q\) if the co-player defected.
The most abundant conditional cooperators in each simulation differ as a
result of the updating payoffs. More specifically, in order for a resident
population of conditional cooperators to avoid being invaded they need to adopt
a different value of \(q\). For each method we have discussed this under the
invasion analysis subsection.

In the cases of perfect memory the resident population adopts a \(q \leq 1 -
\frac{c}{b}=0.9\). In the limited memory case the generosity of a conditional
cooperator is independently of the benefit lower than \(\frac{1}{2}\). The rest
of the cases also condition on the cost benefit relationship. In these
simulations the cost of cooperation is set to 1 and the benefit to 10. As a
result, in the case of two interactions \(q \leq \frac{\delta - 1 +
\frac{\sqrt{2}}{2}}{\delta}=0.7068\), in the case of two rounds \(q \leq
\frac{\delta + \sqrt{\delta^{2} + 1} - 1}{2 \delta}=0.7069\), and in the last
case \(q \leq 1 - \frac{c}{b}=0.9\). The higher tolerance to defection results
to a more cooperative population. As a result the expected payoffs allow for the
most cooperative population. Between the limited memory approaches we observe a
big difference in the cases of one or more information. We hypothesis that as we
allow for more information the results will tend to the case of perfect memory.

\begin{figure}[!htbp]
    \centering 
    \includegraphics[width=\textwidth]{static/more_memory_heatmaps_donation_game_with_illustrations.pdf}
    \caption{\textbf{Evolutionary dynamics with difference updating approach.}
    From left to right, we present result on the following updating payoffs
    cases; the expected payoffs (perfect memory), the last round payoff
    from one interaction (limited memory), the last round payoff from two
    interactions, the last two rounds payoffs from one interaction, the
    last two rounds payoffs from two interactions. We run each simulation for \(T
    = 10^7\) time steps. For each time step we recorded the current resident
    population \((y, p, q)\). Since \(\delta \rightarrow 1\) we do not report the
    players' initial cooperation probability \(y\). The graphs show how often the
    resident population chooses each combination \((p, q)\) of conditional
    cooperation probabilities in the subsequent rounds. In both cases players
    update based on their expected payoffs.}\label{fig:expected_payoffs_results}
\end{figure}

\section{Expected and Last Round Updating Payoffs for Memory One Strategies}\label{section:memory_one}

So far we have assumed that individuals can adopt reactive strategies.
To demonstrate that our results hold for higher memory strategies here we
present results for the expected payoffs, and the last round payoff when members
use memory-one strategies. Memory-one strategies consider the outcome of the
previous round to decide on an action. There are four possible outcome in each
round; \((C, C), (C, D), (D, C), (D, D)\). A memory-one strategy \(s\) can be
written as a five-dimensional vector \(s=(y, p_1, p_2, p_3, p_4)\). The
parameter \(y\) is the probability that the strategy opens with a cooperation
and \(p_1\), \(p_2\), \(p_3\), \(p_4\) are the probabilities that the strategy
cooperates for each of the possible outcomes of the last round.

We perform four separate simulations where we differ the updating payoff and the
benefit of cooperation \(b\). The results for a low value of benefit are given
in Figure~\ref{fig:memory_one_low_benefit}, and for a high benefit in
Figure~\ref{fig:memory_one_high_benefit}. We verify that even when individuals
are allowed to use memory-one strategies, the cooperation rate is higher
in the perfect memory approach compared to the limited memory.

@@

In this section we evaluate the main result of this work for \(\mu \neq 0\).
Namely, we explore the evolved population when individuals use perfect and
limited updating payoff memory for different values of \(\mu\). We perform five
independent runs of the pairwise process described in
Section~\ref{section:pairwise_comparison}, and at each time step we record the
average player \(\bar{s}=(\bar{y}, \bar{p}, \bar{q})\). The average cooperation
of the resident population for different values of mutation are shown in
Figure~\ref{fig:mutation}. The cooperation rate in the case of perfect
memory is always higher compared to the limited memory regardless of the
mutation value. For mutation value of 1 the processes become random and
this results to a cooperation rate of \(\frac{1}{2}\) in both simulations.
